---
title: "Predicting type of activity"
author: "Priya Kasi-Chari"
date: "March 20, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – 
a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a 
particular activity they do, but they rarely quantify how well they do it. This project will  use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise.

Based on the below findings, the predictions are as follows:


## Loading Data & Install Libraries

The following libraries were installed to be used in this project.

```{r}
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(randomForest)
library(knitr)
```

Data from the provided URLs will be loaded into the training and test data sets. A seed is set to enable reproducability. 

```{r}
set.seed(12345)
dataurl_train <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
dataurl_test <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

ds_train <- read.csv(url(dataurl_train), na.strings=c("NA","#DIV/0!",""))
ds_test <- read.csv(url(dataurl_test), na.strings = c("NA","#DIV/0!",""))
```

Since the test dataset provided only has 20 records, the training dataset is split into two - 60% for training the prediction and 40% for validating the model used.

```{r}
inTrain <- createDataPartition(ds_train$classe, p=0.6, list=FALSE)
newTraining <- ds_train[inTrain, ]
newTesting <- ds_train[-inTrain, ]

dim(newTraining)
dim(newTesting)
```

## Cleaning Data

Columns with near zero variance are removed.

```{r}
nearzero <- nearZeroVar(newTraining, saveMetrics = TRUE)
newTraining <- newTraining[,nearzero$nzv==FALSE]

nearzero1 <- nearZeroVar(newTesting, saveMetrics = TRUE)
newTesting <- newTesting[,nearzero1$nzv==FALSE]
```

The first column from newTraining is removed.

```{r}
newTraining <- newTraining[c(-1)]
```


Columns with more than 70% NA values are cleaned in newTraining.

```{r}
temp_column <- newTraining

for (i in 1:length(newTraining)) {
    if( sum( is.na( newTraining[, i] ) ) / nrow(newTraining) >= .7) {
        for(j in 1:length(temp_column)) {
            if( length( grep(names(newTraining[i]), names(temp_column)[j]) ) == 1)  {
                temp_column <- temp_column[ , -j]
            }
        }
    }
}
```

Original column names are replaced in newTraining.

```{r}
newTraining <- temp_column
rm(temp_column)
dim(newTraining)
```

newTesting and ds_test are transformed to include the same columns as newTraining.

```{r}
temp1 <- colnames(newTraining)
temp2 <- colnames(newTraining[,-58])
newTesting <- newTesting[temp1]
ds_test <- ds_test[temp2]

dim(newTesting)
dim(ds_test)
```

All data is coerced to the same type in ds_test.

```{r}
for (i in 1:length(ds_test)) {
    for (j in 1:length(newTraining)) {
        if( length( grep(names(newTraining[i]), names(ds_test)[j]) ) == 1) {
            class(ds_test[j]) <- class(newTraining[i])
        }
    }
}
ds_test <- rbind(newTraining[2, -58], ds_test)
ds_test <- ds_test[-1,]

```

## Predictions with multiple models

The following models will be used to find and predict 'classe' in ds_test with greatest accuracy: Decision Trees, Random Forests and Generalized Boosting.

### Prediction with Random Forests

```{r}
set.seed(12345)
RFFit1 <- randomForest(classe ~., data=newTraining)
RFPrediction1 <- predict(RFFit1, newTesting, type = "class")
RFconfMtx1 <- confusionMatrix(RFPrediction1, newTesting$classe)
RFconfMtx1
```

```{r}
plot(RFFit1)
```

### Prediction with Decision Tree

```{r}
set.seed(12345)
DTFit2 <- rpart(classe ~., data = newTraining, method = "class")
fancyRpartPlot(DTFit2)
```

```{r}
DTPrediction2 <- predict(DTFit2, newTesting, type = "class")
DTconfMtx2 <- confusionMatrix(DTPrediction2, newTesting$classe)
DTconfMtx2
```

### Prediction with Generalized Boosting Model

```{r}
set.seed(12345)
fitsetting <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
gbmFit3 <- train(classe ~ ., data = newTraining, method = "gbm", trControl = fitsetting, verbose = FALSE)
gbmFinMod <- gbmFit3$finalModel
gbmPrediction3 <- predict(gbmFit3, newdata = newTesting)
gbmconfMtx3 <- confusionMatrix(gbmPrediction3, newTesting$classe)
gbmconfMtx3
```

